Single-seismometer Focal Mechanism and Uncertainty Estimation from Body Waves

### 2024-11-05 paper/main.tex
.
Finished bulk of clean codebase + heavy-duty algos

% import packages
\usepackage{amssymb, nicefrac, bm, tikz}

@@

    Our goal is two-fold:
    \begin{enumerate}
      \item [i)]
        Find estimates $\widehat{\psi}, \widehat{\delta}, \widehat{\lambda}$ for the strike, dip
        and rake that \textit{best} describe the collected data.

@@

    \begin{align} \label{eq:3}
      \Phi(\bm{p}) = \frac{-\bm{A}_o^\top\bm{A}_s}{\|\bm{A}_o\|\|\bm{A}_s\|}
    \end{align}

    where $\bm{A}_o = (A_o^P, A_o^{SV}, A_o^{SH})^\top$ are observed amplitudes, $\bm{A}_s =
    (A_s^P, A_s^{SV}, A_s^{SH})^\top$ are synthetic amplitudes, and
    $\bm{p} = (\psi, \delta, \lambda)^\top$ are the underlying parameters.

@@

    \begin{align} \label{eq:4}
      \Phi(\bm{p}) = \sqrt{\left( \Delta \frac{A^{P}}{A^{SV}}\right)^2 + \left( \Delta \frac{A^{SV}}{A^{SH}}\right)^2}
    \end{align}

    \begin{align} \label{eq:5}
      \Phi(\bm{p}) = \sqrt{\log^2 \left( \Delta \frac{A^P}{A^{SV}} \right) + \log^2 \left( \Delta \frac{A^{SV}}{A^{SH}} \right)}
    \end{align}

@@

    \begin{align} \label{eq:6}
      &\Phi(\bm{p}) \sim |\Sigma|^{-1/2} \exp \left( - \frac{1}{2} (\bm{A} - \bm{A}_o)^\top \bm{\Sigma}^{-1} (\bm{A} - \bm{A}_o) \right)
    \end{align}

  \subsection{Tolerance derivation}
    Given vectors $\bm{A}_s$ and $\bm{A}_o$, we would like to know the largest cosine
    similarity for which $\bm{A}_s$ is considered an acceptable fit with respect to the recorded amplitudes
    while accounting for the asymmetry of error levels for different amplitudes. Visualized in 3D
    space, we have a confidence ellipsoid centered at $\bm{A}_o$ (equation \ref{eq:7}) such that
    acceptable fits are vectors which fall inside the elliptical cone tangent to it with vertex at the
    origin.
    \begin{align} \label{eq:7}
      E(\bm{p}) = (\bm{A} - \bm{A}_o)^\top \tilde{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = 1
    \end{align}

@@

    \begin{align} \label{eq:12}
      \nonumber \bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} ((1 - \nicefrac{1}{c_1})\bm{A}_o - \bm{A}_o)  = -1\\
      \Rightarrow c_1 = \bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} \bm{A}_o = \|\bm{A}_o^\top, \bm{A}_o\|_{\tilde{\bm{\Sigma}}^{-1}}^2
    \end{align}
    \begin{align} \label{eq:13}
      \nonumber\bm{D} &= (\bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1})^\top \times (\bm{A}_s \times \bm{A}_o)\\
        &= c_1\bm{A}_s - c_2\bm{A}_o
    \end{align}

@@

    Denote $c_3 = \bm{A}_s^\top \tilde{\bm{\Sigma}}^{-1} \bm{A}_s = \|\bm{A}_s^\top, \bm{A}_s\|_{\tilde{\bm{\Sigma}}^{-1}}^2$. Then by
    substitution we have
    ${\bm{D}^\top \tilde{\bm{\Sigma}}^{-1} \bm{D}} = c_1(c_1c_3 - c_2^2)$ and ${\bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} \bm{D}} = 0$.
    Solving then yields the necessary points (denoted $\bm{A}_b$) as
    \begin{align} \label{eq:15}
      \bm{A}_b = \left(1 - \frac{1}{c_1}\right)\bm{A}_o \pm
        \left(\frac{1}{c_1} \sqrt{\frac{c_1 - 1}{c_1c_3 - c_2^2}}\right)(c_1\bm{A}_s - c_2\bm{A}_o)
    \end{align}

@@

  \subsection{Misfit gradient}
    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable 
    nonlinear function of parameters $\bm{p} = (\psi, \delta, \lambda)^\top$, defined by equation 
    \ref{eq:1}. We can thus apply directed search methods to minimize the following cost function:
    \begin{align} \label{eq:17}
      \Phi(\bm{p}) = -\hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

    To calculate the gradient, first note that $\bm{A}$ is a linear combination of nonlinear functions
    of $\bm{p}$ (equation \ref{eq:2}).

@@

    Define a vector $\bm{\eta} = (s_R, q_R, p_R, q_L, p_L)^\top$. Then we can encode
    the misfit as
    \begin{align} \label{eq:18}
      \Phi(\bm{p}) = \Phi(\bm{A}(\bm{\eta}(\bm{p})))
    \end{align}

@@

    i) From $\bm{p}$ to $\bm{\eta}$, we can use equation \ref{eq:2} to define the Jacobian
    as follows:
    \begin{align*}
      &a = \sin(\lambda)\;\;\;\;\;\;\;\;\;\;\;\;\; b = \cos(2\delta)\;\;\;\;\;\;\;\;\;\;\; c = \cos(\lambda)\\
      &d = \sin(2\delta)\;\;\;\;\;\;\;\;\;\;\;\; e = \sin(\delta)\;\;\;\;\;\;\;\;\;\;\;\; f = \cos(\delta)\\
      &g = \sin(\psi - \phi)\;\;\;\;\;\; h = \cos(\psi - \phi)\;\;\;\;\; p = cos(2(\psi - \phi))\\
      &q = \sin(2(\psi - \phi))
    \end{align*}
    \begin{align} \label{eq:19}
      \nonumber&\bm{J_p}(\bm{\eta}) = \\
      &\begin{bmatrix}
        0         &ab         &0.5cd\\
        abh-cfg   &-2adg-ceh  &cbg-afh\\
        2cep+adq  &cfq-abp    &-aeq-0.5cdp\\
        -cfh-abg  &ceg-2adh   &afg+cbh\\
        adp-2ceq  &abq+cfp    &0.5cdq-aep
      \end{bmatrix}
    \end{align}

    ii) From $\bm{\eta}$ to $A$, we have a linear transformation whose Jacobian is the matrix
    of coefficients from equation \ref{eq:1}:
    \begin{align} \label{eq:20}
      \nonumber&\bm{J_\eta}(\bm{A}) = \\
      &\begin{bmatrix}
        \frac{3\cos^2(i)-1}{\alpha^3}    &\frac{-\sin(2i)}{\alpha^3}    &\frac{-sin^2(i)}{\alpha^3}   &0    &0\\
        \frac{1.5\sin(2j)}{\beta^3}      &\frac{\cos(2j)}{\beta^3}      &\frac{0.5\sin(2j)}{\beta^3}  &0    &0\\
        0         &0           &0        &\frac{\cos(j)}{\beta^3}       &\frac{\sin(j)}{\beta^3}
      \end{bmatrix}
    \end{align}

    iii) From $\bm{A}$ to $\Phi$, we have a scalar function whose Jacobian is the gradient
    transposed.
    \begin{align} \label{eq:21}
      \nonumber\nabla_{\bm{A}}\Phi &= \frac{\|\bm{A}\|\nabla_{\bm{A}}(-\hat{\bm{A}}_o^\top\bm{A}) + 
      \hat{\bm{A}}_o^\top \bm{A} \nabla_{\bm{A}}(\|\bm{A}\|)}{\|\bm{A}\|^2}\\
            &= \frac{1}{\|\bm{A}\|}(\Phi\hat{\bm{A}} - \hat{\bm{A}}_o)
    \end{align}

    Combining \ref{eq:19}, \ref{eq:20} and \ref{eq:21} gives the overall gradient:
    \begin{align} \label{eq:22}
      \nabla_{\bm{p}}\Phi = \bm{J_p}(\bm{\eta})^\top \bm{J_\eta}(\bm{A})^\top \nabla_{\bm{A}}\Phi
    \end{align}

  \subsection{Optimization algorithms}
    Something about gradient-based methods. Try different algorithms, they may produce
    interesting results >> IEMS 450-2.

    \noindent \textbf{TODO:} Create one 3D image with the results of gradient descent from random points in space.

@@

@@

    \subsection{Kagan angle}
      An alternative way of quantifying model deviations.

>>>>

% import packages
\usepackage{amssymb, nicefrac, bm, tikz, subfigure}

@@

    Our goal is two-fold:
    \begin{enumerate}
      \item [i)]
        Find estimates $\widehat{\psi}, \widehat{\delta}, \widehat{\lambda} \in [0^\circ,360^\circ) \times 
        [0^\circ,90^\circ] \times [-180^\circ,180^\circ)$ for the strike, dip and rake that \textit{best} describe
        the collected data.

@@

    \begin{align} \label{eq:3}
      \Phi(\bm{m}) = \frac{-\bm{A}_o^\top\bm{A}_s}{\|\bm{A}_o\|\|\bm{A}_s\|}
    \end{align}

    where $\bm{A}_o = (A_o^P, A_o^{SV}, A_o^{SH})^\top$ are observed amplitudes, $\bm{A}_s =
    (A_s^P, A_s^{SV}, A_s^{SH})^\top$ are synthetic amplitudes, and
    $\bm{m} = (\psi, \delta, \lambda)^\top$ are the underlying parameters.

@@

    \begin{align} \label{eq:4}
      \Phi(\bm{m}) = \sqrt{\left( \Delta \frac{A^{P}}{A^{SV}}\right)^2 + \left( \Delta \frac{A^{SV}}{A^{SH}}\right)^2}
    \end{align}

    \begin{align} \label{eq:5}
      \Phi(\bm{m}) = \sqrt{\log^2 \left( \Delta \frac{A^P}{A^{SV}} \right) + \log^2 \left( \Delta \frac{A^{SV}}{A^{SH}} \right)}
    \end{align}

@@

    \begin{align} \label{eq:6}
      &\Phi(\bm{m}) \sim |\Sigma|^{-1/2} \exp \left( - \frac{1}{2} (\bm{A} - \bm{A}_o)^\top \bm{\Sigma}^{-1} (\bm{A} - \bm{A}_o) \right)
    \end{align}

  \subsection{Tolerance derivation}
    Given vectors $\bm{A}_s$ and $\bm{A}_o$, we would like to know the largest cosine
    similarity for which $\bm{A}_s$ is considered an acceptable fit with respect to the recorded amplitudes
    while accounting for the asymmetry of error levels for different amplitudes. Visualized in 3D
    space, we have a confidence ellipsoid centered at $\bm{A}_o$ (equation \ref{eq:7}) such that
    acceptable fits are vectors which fall inside the elliptical cone tangent to it with vertex at the
    origin.
    \begin{align} \label{eq:7}
      E(\bm{m}) = (\bm{A} - \bm{A}_o)^\top \tilde{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = 1
    \end{align}

@@

    \begin{align} \label{eq:12}
      \nonumber \bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} ((1 - \nicefrac{1}{c_1})\bm{A}_o - \bm{A}_o)  = -1\\
      \Rightarrow c_1 = \bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} \bm{A}_o = \|\bm{A}_o\|_{\tilde{\bm{\Sigma}}^{-1}}^2
    \end{align}
    \begin{align} \label{eq:13}
      \nonumber\bm{D} &= (\bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1})^\top \times (\bm{A}_s \times \bm{A}_o)\\
        &= c_1\bm{A}_s - c_2\bm{A}_o
    \end{align}

@@

    Denote $c_3 = \bm{A}_s^\top \tilde{\bm{\Sigma}}^{-1} \bm{A}_s = \|\bm{A}_s\|_{\tilde{\bm{\Sigma}}^{-1}}^2$. Then by
    substitution we have
    ${\bm{D}^\top \tilde{\bm{\Sigma}}^{-1} \bm{D}} = c_1(c_1c_3 - c_2^2)$ and ${\bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} \bm{D}} = 0$.
    Solving then yields the necessary points (denoted $\bm{A}_b$) as
    \begin{align} \label{eq:15}
      \bm{A}_b = \left(1 - \frac{1}{c_1}\right)\bm{A}_o \pm
        \left(\frac{1}{c_1} \sqrt{\frac{c_1 - 1}{c_1c_3 - c_2^2}}\right)(c_1\bm{A}_s - c_2\bm{A}_o)
    \end{align}

@@

  \subsection{Misfit gradient}
    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable 
    nonlinear function of parameters $\bm{m} = (\psi, \delta, \lambda)^\top$, defined by equation 
    \ref{eq:1}. We can thus apply directed search methods to minimize the following cost function:
    \begin{align} \label{eq:17}
      \Phi(\bm{m}) = -\hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

    To calculate the gradient, first note that $\bm{A}$ is a linear combination of nonlinear functions
    of $\bm{m}$ (equation \ref{eq:2}).

@@

    Define a vector $\bm{\eta} = (s_R, q_R, p_R, q_L, p_L)^\top$. Then we can encode
    the misfit as
    \begin{align} \label{eq:18}
      \Phi(\bm{m}) = \Phi(\bm{A}(\bm{\eta}(\bm{m})))
    \end{align}

@@

    i) From $\bm{m}$ to $\bm{\eta}$, we can use equation \ref{eq:2} to define the Jacobian
    as follows:
    \begin{align*}
      &a = \sin(\lambda)\;\;\;\;\;\;\;\;\;\;\;\;\; b = \cos(2\delta)\;\;\;\;\;\;\;\;\;\;\; c = \cos(\lambda)\\
      &d = \sin(2\delta)\;\;\;\;\;\;\;\;\;\;\;\; e = \sin(\delta)\;\;\;\;\;\;\;\;\;\;\;\; f = \cos(\delta)\\
      &g = \sin(\psi - \phi)\;\;\;\;\;\; h = \cos(\psi - \phi)\;\;\;\;\; p = cos(2(\psi - \phi))\\
      &q = \sin(2(\psi - \phi))
    \end{align*}
    \begin{align} \label{eq:19}
      \nonumber&\bm{J_p}(\bm{\eta}) = \\
      &\begin{bmatrix}
        0         &ab         &0.5cd\\
        abh-cfg   &-2adg-ceh  &cbg-afh\\
        2cep+adq  &cfq-abp    &-aeq-0.5cdp\\
        -cfh-abg  &ceg-2adh   &afg+cbh\\
        adp-2ceq  &abq+cfp    &0.5cdq-aep
      \end{bmatrix}
    \end{align}

    ii) From $\bm{\eta}$ to $A$, we have a linear transformation whose Jacobian is the matrix
    of coefficients from equation \ref{eq:1}:
    \begin{align} \label{eq:20}
      \nonumber&\bm{J_\eta}(\bm{A}) = \\
      &\begin{bmatrix}
        \frac{3\cos^2(i)-1}{\alpha_h^3}    &\frac{-\sin(2i)}{\alpha_h^3}    &\frac{-sin^2(i)}{\alpha_h^3}   &0    &0\\
        \frac{1.5\sin(2j)}{\beta_h^3}      &\frac{\cos(2j)}{\beta_h^3}      &\frac{0.5\sin(2j)}{\beta_h^3}  &0    &0\\
        0         &0           &0        &\frac{\cos(j)}{\beta_h^3}       &\frac{\sin(j)}{\beta_h^3}
      \end{bmatrix}
    \end{align}

    iii) From $\bm{A}$ to $\Phi$, we have a scalar function whose Jacobian is the gradient
    transposed.
    \begin{align} \label{eq:21}
      \nonumber\nabla_{\bm{A}}\Phi &= \frac{\|\bm{A}\|\nabla_{\bm{A}}(-\hat{\bm{A}}_o^\top\bm{A}) - 
      (-\hat{\bm{A}}_o^\top \bm{A}) \nabla_{\bm{A}}(\|\bm{A}\|)}{\|\bm{A}\|^2}\\
            &= \frac{-1}{\|\bm{A}\|}(\hat{\bm{A}}_o + \Phi\hat{\bm{A}})
    \end{align}

    Combining \ref{eq:19}, \ref{eq:20} and \ref{eq:21} gives the overall gradient:
    \begin{align} \label{eq:22}
      \nabla_{\bm{m}}\Phi = \bm{J_p}(\bm{\eta})^\top \bm{J_\eta}(\bm{A})^\top \nabla_{\bm{A}}\Phi
    \end{align}

  \subsection{Optimization algorithms}
    \begin{figure}[ht!]
      \includegraphics[width=8.6cm]{seed_2024} 
      \caption{Results of steepest descent algorithm after a systematic sweep of starting points.}
      \label{fig:2}
    \end{figure}

@@

    \subsection{Kagan angle distribution}
      An alternative way of quantifying model deviations.

@@
